Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-01-18 manuscript/manuscript.Rmd
IACM submission

ggplot(df_exp, aes(x = Performance, y = avg_hardGroup, color = Gender)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),se = FALSE) +
  geom_smooth(data = df_none, aes(x = Performance, y = avg_hardGroup),
              method = "glm", method.args = list(family = "binomial"), color = "grey", se = FALSE) +
  scale_color_manual(values = c("#E74C3C", "#4169E1", "grey"),
                     labels = c("Female + Expectations", "Male + Expectations", "Expectation Hidden")) +
  scale_y_continuous(name = "P(Compete Against High Performance)",
                     labels = scales::percent,
                     expand = c(0, 0)) +
  scale_x_continuous(name = "Performance",
                     expand = c(0, 0),
                     breaks = seq(0, 10, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))
```

@@

@@

To determine whether flattering advice is truly costly, however, we need
to examine the outcome of the advisees. In particular, they could ignore flattering
advice, recognizing it as such and thus failing to adhere to it. Align with our prediction, participants who were primed with high expectations earned less when their expectations were conveyed to the advisers, although this result is only directional  (see Column 1 of Table \@ref(tab:study1time3)). Similarly, as shown in Column 2 of Table \@ref(tab:study1time3), Similarly, as shown in Column 2 of Table 1, we also found that male participants received less reward when their expectations were presented, even when their performance was the same. These findings suggest that flattering advice is not without consequences. Furthermore, our research indicates that because males are more likely to follow such advice (see Column 4 of Table \@ref(tab:study1time3), their tendency towards competition intensifies the cost of such advice.

<!-- David: Let's turn the regression you discussed below into a column of the Table above --> 
<!-- David: Also, one more suggestion for an analysis: let's look at whether Advisees followed the recommendation they received as an other linear probability model (1/0). I would suggest predictors of whether it's the High Performer group, whether the adviser saw expectations, the gender of the advisee. Also, another model that has the two-way interaction for all of those. We may not end up reporting this, but I would be curious to see. -->

<!-- Amanda: Can we create a new table for analyzing time 3 data? Because the variables are different from the previous table, the current table might be difficult to read with new columns. Additionally, the new analysis is only focused on fr advice shown to its target, not all 10,000 advice. It might be confusing if we put them together. -->

<!-- Yes, new table is fine. -->

## Discussion
When the quality of a decision is not dependent on someone's expectations,
advice should not be moved by whether an adviser is aware of the advisees
expectations. However, consistent with our account that advisers consider the
belief utility of the advisee and want to flatter, rather than disappoint them,
we find that displaying expectations to advisers does matter.
Specifically, we found that men underestimate their score on a mathematics test
less than do women. When these expectations are displayed to advisers, they
are more likely to tell men to compete against the High Performer group.
Notably, this is bad advice: men whose advisers observed their expectations
got worse advice as a result. <!-- Amanda: Our findings suggest that males earned less than their female counterparts who scored the same as them, although the evidence is only directional. This could be attributed to the fact that males received more flattering advice and were more likely to adopt it, resulting in even worse outcomes.-->

@@

Next, we recruited a new, gender-balanced sample for the role
of advisers. `r scales::comma(dim(study2_time2)[1])`
participants from Prolific ($M_{\text{Age}}$ =
`r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female) started by
providing demographic information, then were matched to a group of the opposite
gender. They then ranked participants from most to least attractive using a
drop-down menu next to each picture. Because of a limitation with the survey software,
participants were able to select the same rank for multiple participants, rather
than rank them uniquely. We remove XX participants who did not follow instructions
and provide a unique ranking.

<!-- David: Note the number above to fill in. -->

@@

```{r study2advice, fig.cap="Advice given to the 7th most attractive participants"}

@@

ggplot(study2_time2, aes(x = Condition, y = Advice, fill = Condition)) +
    geom_violin(trim = FALSE) +
     scale_fill_manual(values = c("#E74C3C", "#4169E1"), labels = c("Accurate", "Likable")) +
  geom_hline(yintercept = 7, col = 'grey', linetype = 'dashed', linewidth = 0.7) +
  scale_y_continuous(name = "Advised Rank", breaks = seq(1, 10)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

We begin by examining the prior beliefs of advisees who uploaded their selfies.
On average, men guessed they ranked `r round(mean(study2_time1[study2_time1$gender == "Male",]$Guess),2)` in their group of 10 and women guessed that they ranked
`r round(mean(study2_time1[study2_time1$gender == "Female",]$Guess),2)`.
Participants overall underestimate their attractiveness relative to the benchmark
average of 5.5. Moreover, women do so more than men
`r papaja::apa_print(t.test(Guess ~ gender, data = study2_time1, var.equal = T))$statistic`.
Notably, people's self-perceptions correlated strongly with the aggregate
ratings of the advisers (`r round(cor(study2_finalRank$Guess, study2_finalRank$finalRank),2)`, `r papaja::apa_print(cor.test(study2_finalRank$Guess, study2_finalRank$finalRank))$statistic`). However, there
was substantial heterogeneity in perceptions of attractiveness.
Of the 200 participants, `r length(unique(study2_time2$Target))` were ranked as 7th most attractive by at least one adviser. 
On average, men in this subset estimated they were ranked `r round(mean(TargetsUnique[TargetsUnique$gender == 'Male',]$Guess),2)`th and women estimated they were
ranked `r `round(mean(TargetsUnique[TargetsUnique$gender == 'Female',]$Guess),2)`th (`r papaja::apa_print(t.test(Guess ~ gender, data = TargetsUnique, var.equal = T))$statistic`).

@@

```{r study2Rating, fig.cap="The evaluation of advice givers", eval = F}
# Load the necessary packages
library(ggplot2)
library(gridExtra)
library(dplyr)

@@

# Create the first plot
plot1 <- ggplot(study2_time3, aes(x = Advice)) +
  geom_boxplot(aes(y = likable),position = position_dodge()) +
  scale_y_continuous(name = "Rating on Likability") +
  scale_x_discrete(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

# Create the second plot
plot2 <- ggplot(study2_time3, aes(x = Advice)) +
  geom_boxplot(aes(y = warmAvg),position = position_dodge()) +
  scale_y_continuous(name = "Rating on Warmth") +
  scale_x_discrete(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

Finally, we examine whether flattering advice indeed leads to more positive evaluations of
advisers, or whether flattering advice is dismissed as insencere.
Following our preregistration, we average the ratings on likeability,
warmth,  friendliness, and good-naturedness to create a scale of likeability
($\alpha$ = `r cronbach.alpha(study2_time3[, 5:8])$alpha`);
<!-- David: The draft had trustworthiness as part of this scale. -->
<!-- Can we just double-check against the preregistration? -->
<!-- And was Likeable part of that? -->
<!-- We might also want a name for the scale that isn't the same as one -->
<!-- of the items, though I'm not sure what. -->
<!-- Same for trustworthiness -- though that could simply be "Trust" -->

<!-- Amanda: My bad. We only pre-reg that we will measure likable (single item) and warm (four items)" -->
<!-- I include trustworthiness here because it is part of that existing scale" -->

and we create a scale of trustworthiness by averaging the ratings of
trustworthiness and sincerity ($\alpha$ = `r cronbach.alpha(study2_time3[, 9:10])$alpha`).

<!-- Amanda: How about the following picture? -->

```{r study2ratingByGroup, fig.cap="The Response to Different Advice"}

@@

ggplot(rating_df, aes(x = Advice, y = Freq, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("#E74C3C",'#4169E1')) +
  labs(x = "Advice", y = "Rating") +
  ggtitle('Response to Advice') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20)) +
  geom_text(aes(label = paste0(Freq), y = 0.5), position = position_dodge(width = 0.9), size = 2, color = 'white')

@@

As seen in Figure \@ref(fig:study2Rating), we
found that advisees who suggested that the advisee was more attractive
(lower rank) were indeed rated as more likeable and warm (b = -0.109, p
\< .05; b = -0.133, p \< .001, respectively; See Columns 1 and 2 of Table \@ref(tab:study2regs)).
Interestingly, these benefits are not at the cost of being hypocritical;
advisers who recommend a more favorable rank are viewed as no less trustworthy
(Column 3 of Table \@ref(tab:study2regs)).

>>>>

ggplot(df_exp, aes(x = Performance, y = avg_hardGroup, color = Gender)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),se = FALSE) +
  geom_smooth(data = df_none, aes(x = Performance, y = avg_hardGroup),
              method = "glm", method.args = list(family = "binomial"), color = "grey", se = FALSE) +
  scale_color_manual(values = c("firebrick3",'dodgerblue3', "grey"),
                     labels = c("Female + Expectations", "Male + Expectations", "Expectation Hidden")) +
  scale_y_continuous(name = "P(Compete Against High Performance)",
                     labels = scales::percent,
                     expand = c(0, 0)) +
  scale_x_continuous(name = "Performance",
                     expand = c(0, 0),
                     breaks = seq(0, 10, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))
```

@@

<!-- Amanda: I think our primary purpose here is still showing the existence of gender difference -->
<!-- The current plot shows discrepancy between two lines, allowing us to demonstrate gender gap -->

@@

To determine whether flattering advice is truly costly, however, we need
to examine the outcome of the advisees. In particular, they could ignore flattering
advice, recognizing it as such and thus failing to adhere to it. Align with our prediction, participants who were primed with high expectations earned less when their expectations were conveyed to the advisers, although this result is only directional  (see Column 1 of Table \@ref(tab:study1time3)). Similarly, as shown in Column 2 of Table \@ref(tab:study1time3), we also found that male participants received less reward when their expectations were presented, even when their performance was the same. These findings suggest that flattering advice is not without consequences. Furthermore, our research indicates that because males are more likely to follow such advice (see Column 4 of Table \@ref(tab:study1time3), their tendency towards competition intensifies the cost of such advice.

## Discussion
When the quality of a decision is not dependent on someone's expectations,
advice should not be moved by whether an adviser is aware of the advisees
expectations. However, consistent with our account that advisers consider the
belief utility of the advisee and want to flatter, rather than disappoint them,
we find that displaying expectations to advisers does matter.
Specifically, we found that men underestimate their score on a mathematics test
less than do women. When these expectations are displayed to advisers, they
are more likely to tell men to compete against the High Performer group.
Notably, this is bad advice: men whose advisers observed their expectations
got worse advice as a result. Our findings suggest that males with the same scores as their female counterparts earned less, although the evidence is only directional. This may be due to males receiving more favorable advice and being more likely to adopt it, resulting in even worse outcomes.

@@

Next, we recruited a new, gender-balanced sample for the role
of advisers. `r scales::comma(dim(study2_time2)[1])`
participants from Prolific ($M_{\text{Age}}$ =
`r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female) started by
providing demographic information, then were matched to a group of the opposite
gender. They then ranked participants from most to least attractive using a
drop-down menu next to each picture. Because of a limitation with the survey software,
participants were able to select the same rank for multiple participants, rather
than rank them uniquely. We remove 115 participants who did not follow instructions
and provide a unique ranking.

@@

```{r study2advice, fig.cap="Advisers were asked to suggest what rank to bet on for someone they ranked as the 7th most attractive out of ten people. When incentivized for accuracy, most of them advised betting on 7, which was based on their evaluation of the recipient's attractiveness. However, when incentivized for likability, many participants gave flattering advice by suggesting betting on higher ranks that imply a higher evaluation of attractiveness."}

@@

ggplot(study2_time2, aes(x = Condition, y = Advice, fill = Condition)) +
    geom_violin(trim = FALSE) +
     scale_fill_manual(values = c("firebrick3",'dodgerblue3'), labels = c("Accurate", "Likable")) +
  geom_hline(yintercept = 7, col = 'grey', linetype = 'dashed', linewidth = 0.7) +
  scale_y_continuous(name = "Advised Rank", breaks = seq(1, 10)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

We begin by examining the prior beliefs of advisees who uploaded their selfies.
On average, men guessed they ranked `r round(mean(study2_time1[study2_time1$gender == "Male",]$Guess),2)` in their group of 10 and women guessed that they ranked
`r round(mean(study2_time1[study2_time1$gender == "Female",]$Guess),2)`.
Participants overall underestimate their attractiveness relative to the benchmark
average of 5.5. Moreover, women do so more than men
`r papaja::apa_print(t.test(Guess ~ gender, data = study2_time1, var.equal = T))$statistic`.
Notably, people's self-perceptions correlated strongly with the aggregate
ratings of the advisers (`r round(cor(study2_finalRank$Guess, study2_finalRank$finalRank),2)`, `r papaja::apa_print(cor.test(study2_finalRank$Guess, study2_finalRank$finalRank))$statistic`). However, there
was substantial heterogeneity in perceptions of attractiveness.
Of the 200 participants, `r length(unique(study2_time2$Target))` were ranked as 7th most attractive by at least one adviser. 
On average, men in this subset estimated they were ranked `r round(mean(TargetsUnique[TargetsUnique$gender == 'Male',]$Guess),2)`th and women estimated they were
ranked `r round(mean(TargetsUnique[TargetsUnique$gender == 'Female',]$Guess),2)`th (`r papaja::apa_print(t.test(Guess ~ gender, data = TargetsUnique, var.equal = T))$statistic`).

@@

Finally, we examine whether flattering advice indeed leads to more positive evaluations of
advisers, or whether flattering advice is dismissed as insencere.
Following our preregistration, we average the ratings on likeability,
warmth,  friendliness, and good-naturedness to create a scale of likeability
($\alpha$ = `r cronbach.alpha(study2_time3[, 5:8])$alpha`); and we create a scale of trustworthiness by averaging the ratings of trustworthiness and sincerity ($\alpha$ = `r cronbach.alpha(study2_time3[, 9:10])$alpha`).

<!-- Amanda: How about the following picture? -->
```{r study2rating, fig.cap="Participants tend to rate advisors who suggest betting on a higher rank (implying greater attractiveness) as more likable and warm."}

library(ggplot2)
library(gridExtra)
library(dplyr)

@@

# Create the first plot
plot1 <- ggplot(study2_time3, aes(x = Advice)) +
  geom_boxplot(aes(y = likable),position = position_dodge(width = 0.8), alpha = 0.8, color = "black", fill = "firebrick3") +
  scale_y_continuous(name = "Likability") +
  scale_x_discrete(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

# Create the second plot
plot2 <- ggplot(study2_time3, aes(x = Advice)) +
  geom_boxplot(aes(y = warmAvg),position = position_dodge(width = 0.8), alpha = 0.8, color = "black", fill = "dodgerblue3") +
  scale_y_continuous(name = "Warmth") +
  scale_x_discrete(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

```{r study2ratingByGroup, fig.cap="After participants received advice that suggested betting on a rank that was lower than, equal to, or higher than their self-expectations, they evaluated advisors based on their likability and warmth. Participants rated advisors who gave flattering advice and delivered a high evaluation of their attractiveness as more likable and warm than those whose advice implied a lower evaluation of their attractiveness."}

@@

ggplot(rating_df, aes(x = Advice, y = Freq, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("firebrick3",'dodgerblue3')) +
  labs(x = "Advice", y = "Rating") +
  ggtitle('Response to Advice') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20)) +
  geom_text(aes(label = paste0(Freq), y = 0.5), position = position_dodge(width = 0.9), size = 5, color = 'white', fontface = "bold")

@@

As seen in Figure \@ref(fig:study2rating), we
found that advisees who suggested that the advisee was more attractive
(lower rank) were indeed rated as more likeable and warm (b = -0.109, p
\< .05; b = -0.133, p \< .001, respectively; See Columns 1 and 2 of Table \@ref(tab:study2regs)). When using their self-expectations as a reference point, advisors who suggest a rank lower than that point are perceived as less likable and warm (see Figure \@ref(fig:study2ratingByGroup)).
Interestingly, these benefits are not at the cost of being hypocritical;
advisers who recommend a more favorable rank are viewed as no less trustworthy
(Column 3 of Table \@ref(tab:study2regs)).
